The evolution of the transcriptome of mammalian placentas

### 2016-09-08 manuscript/extended_discussion.Rnw
copy manuscript to extended discussion

\title{The core transcriptome of the mammal placenta}%
%
\author[affigb]{Don L. Armstrong\fnref{n1}}
\ead{don@donarmstrong.com}
\author[afflon]{Michael R. McGowen\fnref{n1}}
\ead{1mmcgowen1@gmail.com}
\author[affigb]{Amy Weckle}
\ead{aweckle@illinois.edu}
\author[affwayne]{Jason Caravas}
\ead{jacaravas@gmail.com}
\author[affmich]{Dalen Agnew}
\ead{agnewd@dcpah.msu.edu}
\author[affuscd]{Kurt Benirschke}
\ead{kbenirschke@ucsd.edu}
\author[affapetrust]{Sue Savage-Rumbaugh}
\ead{suerumbaugh@gmail.com}
\author[affhaifa]{Eviatar Nevo}
\ead{nevo@research.haifa.ac.il}
\author[affnih,affwaynepath]{Chong J Kim}
\ead{cjkim@med.wayne.edu}
\author[affyale]{Günter P. Wagner}
\ead{gunter.wagner@yale.edu}
\author[affnih]{Roberto Romero}
\ead{prbchiefstaff@med.wayne.edu}
\author[affigb,affapetrust,affmip]{Derek E. Wildman\corref{cor1}}
\ead{wildmand@illinois.edu}
\cortext[cor1]{Corresponding author}
\address[affigb]{Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
\address[afflon]{School of Biological and Chemical Sciences, Queen Mary, University of London, London, UK}
\address[affwayne]{Center for Molecular Medicine and Genetics, Wayne State University School of Medicine, Detroit, MI, USA}
\address[affmich]{Diagnostic Center for Population and Animal Health, College of Veterinary Medicine, Michigan State University, East Lansing, MI, USA}
\address[affuscd]{Department of Pathology, University of California San Diego, La Jolla, CA, USA}
\address[affapetrust]{Great Ape Trust/Bonobo Hope Sanctuary, Des Moines, IA, USA}
\address[affnih]{Perinatology Research Branch, Eunice Kennedy Shriver National Institute of Child Health and Human Development (NICHD), National Institutes of Health (NIH), Detroit, MI, USA}
\address[affwaynepath]{Department of Pathology, Wayne State University School of Medicine, Detroit, MI, USA}
\address[affyale]{Department of Ecology and Evolutionary Biology and Yale Systems Biology Institute, Yale University, New Haven, CT, USA}
\address[affmip]{Department of Molecular and Integrative Physiology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
\address[affhaifa]{The Institute of Evolution, University of Haifa, Haifa, Israel}

@@

Supplemental \cref{fig:transcriptome_statistics} shows the RNAseq read and
alignment statistics for all species. For transcriptomes sequenced
with 75 bp %
%
(\Sexpr{species(star.logs.wide[input.read.length > 70 & input.read.length < 80,unique(species)])}),%
%
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,max(as.numeric(input.reads))]}
with an average of \Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,mean(`% good reads`)]}\% of reads 
mapping to the respective reference genome (\cref{tab:ref_genome_aligned_species}). 
%
For transcriptomes sequenced with 150 bp %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length > 120 & input.read.length < 160,unique(species)])}),
%
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  > 120 & input.read.length < 160,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,mean(`% good reads`)]}\% %
of reads mapping to the
respective reference genome (\cref{tab:ref_genome_aligned_species}).
%
%
For transcriptomes sequenced with 35 bp %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length < 40 ,unique(species)])}),
%
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length < 40 ,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  < 40  ,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length < 40, mean(`% good reads`)]}\%
of reads mapping to the
respective reference genome (\cref{tab:ref_genome_aligned_species}).
The top 50 expressed genes in each genome analyzed are shown in
\cref{tab:top50_human_placenta,tab:top50_bonobo_placenta,tab:top50_spider_monkey_placenta,tab:top50_dog_placenta,tab:top50_cow_placenta,tab:top50_elephant_placenta,tab:top50_opossum_placenta,tab:top50_mouse_placenta,tab:top50_n_galili_placenta,tab:top50_s_carmeli_placenta,tab:top50_pig_placenta,tab:top50_horse_placenta,tab:top50_sheep_placenta}
and all genes in Supplemental \cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_spalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}.

% The percentage of bases that mapped to Ensembl annotated mRNA ranged
% from ~35\% in elephant to ~85\% in human, with percentages of bases
% mapping to annotated protein-coding segments of genes ranging from
% ~34\% (elephant) to ~54\% (human), untranslated regions (UTRs) ranging
% from ~0.6\% (elephant) to ~31\% (human), and known introns ranging
% from ~5.5\% (dog) to 11\% (bonobo). Median percent genomic coverage
% ranged from ~43\% in elephant to 80\% in bonobo. The number of
% annotated transcripts to which reads mapped in each respective genomes
% ranged from 4119 (bonobo) to 6506 (human) for transcripts with \ac{FPKM}>10
% and 11689 (dog) to 30733 (human) for transcripts with \ac{FPKM}>1. The
% greater degree of mapping to annotated protein-coding sequences in
% humans corresponds to the greater number of Ensembl annotated
% sequences in the human genome compared to the other five reference
% genomes (Flicek, 2012).

@@

<<gene.expression,results="hide",echo=FALSE>>=
mentioned.genes <- load_cache(label="save_mentioned_genes",object="saved_mentioned_genes",list())
gene.expression <- function(species.name,gene,complete=FALSE,units=TRUE) {
    species.name <- gsub("_"," ",species.name)
    mentioned.genes[[paste0(species.name,"_",gene)]] <<-
        list("species.name"=species.name,
             "gene"=gene)
    fpkm <- combined.fpkm[species==species.name &
                              (gene_short_name==gene | 
                               gene_id==gene),
                          mean_fpkm][1]
    if (is.na(fpkm)) {
        stop("FPKM is NA!!!!")
    }
    if (units || complete)
        fpkm <- formatC(fpkm,digits=0,format="f")
    if (units) {
        fpkm <- paste0(fpkm,
                       "\\ \\text{FPKM}")
    }
    if (complete) {
        if (any(grepl("^ENS.+G\\d+$",gene))) {
            fpkm <- paste0("\\text{",gene,"} = ",
                           fpkm)
        } else {
            fpkm <- paste0("\\text{\\textit{",gene,"}} = ",
                           fpkm)
        }
    }
    if (units || complete) {
        fpkm <- paste0("$",fpkm,"$")
    }
    return(fpkm)
}

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \human are shown in
\cref{tab:top50_human_placenta} with all genes with \ac{FPKM} $\ge 50$
in Supplemental \cref{tab:all_homo_sapiens_placenta}. The gene with by far
the highest \ac{FPKM} value
%
(\Sexpr{gene.expression(species.name="homo sapiens",gene="CSH1")}) %
%
in \human is \textit{CSH1}, followed by \textit{CSH2}, and \textit{CSHL1}, all
three of which are members of a distinct expansion of the growth
hormone gene family within catarrhines (Supplemental \cref{fig:galectins_evolution_tree}). Another member of the growth
hormone gene family, \ac{GH2} (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="GH2")}), %
is also expressed to a high degree in the \human placenta
(\cref{tab:top50_human_placenta}).
% 
Eleven \acp{PSG} are highly-expressed
(\ac{FPKM}$\ge 100$) in the term placenta of \human, with \textit{PSG1} showing the
greatest degree of expression %
(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG1")}) % 
and the pseudogene \textit{PSG10P} showing the least %
(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG10P")}, \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~A) . %
These
genes are members of a hominoid-specific expansion of the \ac{CEA} family
and known to be expressed by the syncytiotrophoblast
\cite{Rawn.ea2008:evolutionregulationandfunction,Lei.ea1992:Cloningandexpressionof}. %
%
Genes involved in the production of estrogens (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP19A1",complete=TRUE)},
% 
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD17B1",complete=TRUE)}, and
% 
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)}%
% 
) and progesterone (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)},
\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP11A1",complete=TRUE)}, and
\Sexpr{gene.expression(species.name="homo sapiens",gene="ADM",complete=TRUE)})
%
show exceptionally high expression in the human placenta, as well as %
%
\textit{CGA} (FPKM=\Sexpr{gene.expression(species.name="homo sapiens",gene="CGA")}, %
%
which generates the alpha subnit of multiple human chorionic
gonadotropin hormones (\cref{tab:top50_human_placenta}). The genes
producing the beta subunits of the chorionic gonadotropins
(\textit{CGB}, \textit{CGB1}, \textit{CGB2}, \textit{CGB5},
\textit{CGB7}, \textit{CGB8}) have reasonable expression
($\text{FPKM} \ge \Sexpr{cgb.min.expression}$), and the sum of their
expression is on the order of the expression of
\textit{CGA} ($\sum \mathit{CGBs}=\Sexpr{cgb.sum.expression}$ FPKM) .
%
%
Other highly expressed genes in the top 50 include members of the
hemoglobin family (e.g. \textit{HBA1}, \textit{HBA2}, \textit{HBB},
and \textit{HBG2}), the maternally imprinted long noncoding RNA gene
\textit{H19}, genes coding for proteins required for growth and/or
cell proliferation (\textit{ADAM12}, \textit{KISS1}, \textit{IGF2},
\textit{GPC3}, \textit{CSF3R}), anticoagulation (\textit{TFPI2}), and
glucose transport (\textit{SLC2A1}) among others
(\cref{tab:top50_human_placenta}, Supplemental
\cref{tab:all_homo_sapiens_placenta}).

@@

\begin{figure*}[htb]
<<psg.ctsl.evolution.tree,echo=FALSE,out.width="\\textwidth",fig.width=8,fig.height=8>>=
pushViewport(viewport(layout=grid.layout(nrow=2,ncol=2),clip="off"))
print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="PSG1" & species=="homo sapiens"),
                                genes.to.tree=genes.to.tree,
                                combined.fpkm=combined.fpkm,
                                subtree=c("ENSG00000221826","ENSG00000124469"),
                                trees=trees
                                )+
      geom_tiplab(size=2,color="black",fontface="italic")+xlim(0,0.2)+
      theme(legend.position="none"),
      vp=viewport(layout.pos.col=1,layout.pos.row=1,clip="off"))
print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="CTSL" & species=="homo sapiens"),
                                genes.to.tree=genes.to.tree,
                                combined.fpkm=combined.fpkm,
                                subtree=c("ENSMODG00000002844","ENSMUSG00000055679"),
                                trees=trees)+
      geom_tiplab(size=2,color="black",fontface="italic")+xlim(0,1)+
      theme(legend.position="none"),
      vp=viewport(layout.pos.col=2,layout.pos.row=1,clip="off"))
print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="PRSS16" &
                                                    species=="homo sapiens"),
                                genes.to.tree=genes.to.tree,
                                subtree=c(   "ENSG00000112812",
                                          "ENSLAFG00000032211"),
                                combined.fpkm=combined.fpkm,
                                trees=trees)+
      geom_tiplab(size=2,color="black",fontface="italic")+xlim(0,1.2)+
      theme(legend.position="none"),
      vp=viewport(layout.pos.col=1,layout.pos.row=2,clip="off"))
print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="PAG2" &
                                                    species=="bos taurus"),
                                genes.to.tree=genes.to.tree,
                                combined.fpkm=combined.fpkm,
                                subtree=c("ENSSSCG00000013097",
                                          "ENSBTAG00000046085"),
                                trees=trees) +
      geom_tiplab(size=2,color="black",fontface="italic") +
      xlim(0,0.8),
      vp=viewport(layout.pos.col=2,layout.pos.row=2,clip="off"))
popViewport(1)

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \bonobo are shown in \cref{tab:top50_bonobo_placenta}
with a list of all genes in Supplemental
\cref{tab:all_pan_paniscus_placenta}. Expression in the \bonobo
placenta differs somewhat from that of \human, as evidenced by the
makeup of the top 50 genes (\cref{tab:top50_bonobo_placenta}). The highest expression
 in \bonobo is %
\Sexpr{gene.expression("pan paniscus","PRG2",complete=TRUE)}, %
%
which is involved in insulin growth factor signaling and angiogenesis
during pregnancy \cite{Weyer.ea2011:Placentalregulationofpeptide}.
\textit{PRG2} is also expressed at relatively high levels (though
comparatively much lower) in the human placenta
(\Sexpr{gene.expression("homo sapiens","PRG2")};
Supplemental \cref{tab:all_homo_sapiens_placenta}). Four \bonobo genes
related to the chorionic somatomammotropins
are also expressed at relatively high levels (\textit{GH1} 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034560",complete=TRUE)}, 
\textit{GH1} 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034212",complete=TRUE)}, 
\textit{GH2}
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000030805",complete=TRUE)}, 
and \textit{GH2}
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034209",complete=TRUE)}, 
%
\cref{tab:all_pan_paniscus_placenta}, \cref{fig:csh1_evolution_tree}).
Placental expression of the genes involved in estrogen and
progesterone production (\textit{CYP19A1}, \textit{HSD3B1}
respectively) show lower expression in \bonobo as compared to \human %
(\Sexpr{gene.expression("pan paniscus","CYP19A1",complete=TRUE)},
\cref{fig:estrogen_evolution_tree}; %
\Sexpr{gene.expression("pan paniscus","HSD3B1",complete=TRUE)},
\cref{fig:progesterone_evolution_tree} and
\cref{tab:top50_human_placenta}) although the expression of
\textit{HSD3B1} is fairly similar to that of \smonkey %
(\Sexpr{gene.expression("ateles fusciceps","HSD3B1",complete=TRUE)}. %
In addition, \textit{CGA} is expressed in the bonobo placenta %
(\Sexpr{gene.expression("pan paniscus","CGA")}), but at much
lower levels than \human as well (\cref{tab:top50_human_placenta}).
%
Two Ensembl-annotated beta subunits of the chorionic gonadotropin
genes are expressed in \bonobo, one related to \textit{CGB} and
another related to \textit{CGB7}; both are not highly expressed
(\cref{fig:cga_cgb_evolution_tree}). Among the other top 50 expressed
genes, there is high expression of keratins (e.g., \textit{KRT19},
\textit{KRT18}), heat-shock proteins (e.g., \textit{HSP90B1},
\textit{HSPA1A}), serpins (\textit{SERPINE1}, \textit{SERPINE2},
Supplemental \cref{fig:serpine_evolution_tree}), insulin growth factor
2 (\textit{IGF2}) and one of its binding proteins \textit{IGFBP3}, and
the hormone adrenomedullin (\textit{ADM}).

@@

The \opossum placenta shows expression of genes that either do not
exist in eutherians or are not expressed in the term eutherian placenta
(\cref{tab:top50_opossum_placenta}). The existence of embryonic forms
of hemoglobin in fetal tissues at term, as evidenced by the expression
of \textit{HBE} and \textit{HBZ} is a feature of marsupials that is
generally due to their very short gestation periods. Embryonic
hemoglobins in marsupials have a lower affinity for oxygen and allow
for a lower rate of oxygen uptake from maternal tissues
\cite{Holland.ea1997:Characterizationofembryonicglobin}. One novel
uncharacterized gene that shows high expression (\textit{LOC10001195}) is not
found in eutherians; Ensembl gene trees reveal that homologs of this
gene exist in other amniotes such as birds and turtles, but are not
present in eutherian mammals. The homolog of this gene in chicken is
found to expressed in reproductive tract
\cite{Froman.ea2006:expressedsequencetaganalysis}, but its function is
unknown. \textit{WFDC2} is a member of the whey acidic protein (WAP)
family and involved in innate immunity. It is expressed in the mammary
gland of the tammar wallaby during pregnancy and early lactation
\cite{Sharp.ea2007:Molecularevolutionofmonotreme}, and present in
relative abundance in \opossum placenta.

>>>>

\title{The evolution of the transcriptome of mammalian placentas}%
%
\author[affigb]{Don L. Armstrong\fnref{n1}}
\ead{don@donarmstrong.com}
\author[afflon]{Michael R. McGowen\fnref{n1}}
\ead{1mmcgowen1@gmail.com}
\author[affigb]{Amy Weckle}
\ead{aweckle@illinois.edu}
\author[affwayne]{Jason Caravas}
\ead{jacaravas@gmail.com}
\author[affmich]{Dalen Agnew}
\ead{agnewd@dcpah.msu.edu}
\author[affuscd]{Kurt Benirschke}
\ead{kbenirschke@ucsd.edu}
\author[affapetrust]{Sue Savage-Rumbaugh}
\ead{suerumbaugh@gmail.com}
\author[affhaifa]{Eviatar Nevo}
\ead{nevo@research.haifa.ac.il}
\author[affnih,affwaynepath]{Chong J Kim}
\ead{cjkim@med.wayne.edu}
\author[affyale]{Günter P. Wagner}
\ead{gunter.wagner@yale.edu}
\author[affnih]{Roberto Romero}
\ead{prbchiefstaff@med.wayne.edu}
\author[affigb,affapetrust,affmip]{Derek E. Wildman\corref{cor1}}
\ead{wildmand@illinois.edu}
\cortext[cor1]{Corresponding author}
\address[affigb]{Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
\address[afflon]{School of Biological and Chemical Sciences, Queen Mary, University of London, London, UK}
\address[affwayne]{Center for Molecular Medicine and Genetics, Wayne State University School of Medicine, Detroit, MI, USA}
\address[affmich]{Diagnostic Center for Population and Animal Health, College of Veterinary Medicine, Michigan State University, East Lansing, MI, USA}
\address[affuscd]{Department of Pathology, University of California San Diego, La Jolla, CA, USA}
\address[affapetrust]{Great Ape Trust/Bonobo Hope Sanctuary, Des Moines, IA, USA}
\address[affnih]{Perinatology Research Branch, Eunice Kennedy Shriver National Institute of Child Health and Human Development (NICHD), National Institutes of Health (NIH), Detroit, MI, USA}
\address[affwaynepath]{Department of Pathology, Wayne State University School of Medicine, Detroit, MI, USA}
\address[affyale]{Department of Ecology and Evolutionary Biology and Yale Systems Biology Institute, Yale University, New Haven, CT, USA}
\address[affmip]{Department of Molecular and Integrative Physiology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
\address[affhaifa]{The Institute of Evolution, University of Haifa, Haifa, Israel}

@@

Supplemental \cref{fig:transcriptome_statistics} shows the RNAseq read and
alignment statistics for all species. For transcriptomes sequenced
with 75 bp %
%
(\Sexpr{species(star.logs.wide[input.read.length > 70 & input.read.length < 80,unique(species)])}),%
%
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,max(as.numeric(input.reads))]}
with an average of \Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,mean(`% good reads`)]}\% of reads 
mapping to the respective reference genome (\cref{tab:ref_genome_aligned_species}). 
%
For transcriptomes sequenced with 150 bp %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length > 120 & input.read.length < 160,unique(species)])}),
%
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  > 120 & input.read.length < 160,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,mean(`% good reads`)]}\% %
of reads mapping to the
respective reference genome (\cref{tab:ref_genome_aligned_species}).
%
%
For transcriptomes sequenced with 35 bp %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length < 40 ,unique(species)])}),
%
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length < 40 ,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  < 40  ,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length < 40, mean(`% good reads`)]}\%
of reads mapping to the
respective reference genome (\cref{tab:ref_genome_aligned_species}).
The expression of all genes with \ac{FPKM} $\ge 50$ is shown in Supplemental
\cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_spalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}.

@@

<<gene.expression,results="hide",echo=FALSE>>=
mentioned.genes <- load_cache(label="save_mentioned_genes",object="saved_mentioned_genes",list())
gene.expression <- function(species.name,gene,complete=FALSE,units=TRUE,mention.only=FALSE) {
    species.name <- gsub("_"," ",species.name)
    mentioned.genes[[paste0(species.name,"_",gene)]] <<-
        list("species.name"=species.name,
             "gene"=gene)
    if (mention.only) {
        return("")
    }
    fpkm <- combined.fpkm[species==species.name &
                              (gene_short_name==gene | 
                               gene_id==gene),
                          mean_fpkm][1]
    if (is.na(fpkm)) {
        stop("FPKM is NA!!!!")
    }
    if (units || complete)
        fpkm <- formatC(fpkm,digits=0,format="f")
    if (units) {
        fpkm <- paste0(fpkm,
                       "\\ \\text{FPKM}")
    }
    if (complete) {
        if (any(grepl("^ENS.+G\\d+$",gene))) {
            fpkm <- paste0(gsub("(ENS)(.+)(G)0*(\\d+)","\\1\\\\linebreak[0]\\2\\\\linebreak[0]\\3...\\4",gene)," = ",
                           fpkm)
        } else {
            fpkm <- paste0("\\textit{",gene,"} = ",
                           fpkm)
        }
    }
    ## if (units || complete) {
    ##     fpkm <- paste0("$",fpkm,"$")
    ## }
    return(fpkm)
}

@@

The gene with by far the highest \ac{FPKM} value
%
(\Sexpr{gene.expression(species.name="homo sapiens",gene="CSH1")}) %
%
in \human is \textit{CSH1}, followed by \textit{CSH2}, and \textit{CSHL1}, all
three of which are members of a distinct expansion of the growth
hormone gene family within catarrhines (Supplemental \cref{fig:galectins_evolution_tree}). Another member of the growth
hormone gene family, \ac{GH2} (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="GH2")}), %
is also expressed to a high degree in the \human placenta
(Supplemental \cref{tab:all_homo_sapiens_placenta}).
% 
Eleven \acp{PSG} are highly-expressed
(\ac{FPKM}$\ge 100$) in the term placenta of \human, with \textit{PSG1} showing the
greatest degree of expression %
(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG1")}) % 
and the pseudogene \textit{PSG10P} showing the least %
(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG10P")}, \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~A) . %
These
genes are members of a hominoid-specific expansion of the \ac{CEA} family
and known to be expressed by the syncytiotrophoblast
\cite{Rawn.ea2008:evolutionregulationandfunction,Lei.ea1992:Cloningandexpressionof}. %
%
Genes involved in the production of estrogens (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP19A1",complete=TRUE)},
% 
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD17B1",complete=TRUE)}, and
% 
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)}%
% 
) and progesterone (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)},
\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP11A1",complete=TRUE)}, and
\Sexpr{gene.expression(species.name="homo sapiens",gene="ADM",complete=TRUE)})
%
show exceptionally high expression in the human placenta, as well as %
%
\textit{CGA} (FPKM=\Sexpr{gene.expression(species.name="homo sapiens",gene="CGA")}, %
%
which generates the alpha subnit of multiple human chorionic
gonadotropin hormones (Supplemental \cref{tab:all_homo_sapiens_placenta}). The genes
producing the beta subunits of the chorionic gonadotropins
(\textit{CGB}, \textit{CGB1}, \textit{CGB2}, \textit{CGB5},
\textit{CGB7}, \textit{CGB8}) have reasonable expression
($\text{FPKM} \ge \Sexpr{cgb.min.expression}$), and the sum of their
expression is on the order of the expression of
\textit{CGA} ($\sum \mathit{CGBs}=\Sexpr{cgb.sum.expression}$ FPKM) .
%
%
Other highly expressed genes in the top 50 include members of the
hemoglobin family (e.g. \textit{HBA1}, \textit{HBA2}, \textit{HBB},
and \textit{HBG2}), the maternally imprinted long noncoding RNA gene
\textit{H19}, genes coding for proteins required for growth and/or
cell proliferation (\textit{ADAM12}, \textit{KISS1}, \textit{IGF2},
\textit{GPC3}, \textit{CSF3R}), anticoagulation (\textit{TFPI2}), and
glucose transport (\textit{SLC2A1}) among others
(Supplemental \cref{tab:all_homo_sapiens_placenta}, Supplemental
\cref{tab:all_homo_sapiens_placenta}).

@@

\begin{figure*}[htb]
<<psg.ctsl.evolution.tree,echo=FALSE,out.width="\\textwidth",fig.width=8,fig.height=8>>=
pushViewport(viewport(layout=grid.layout(nrow=2,ncol=2),clip="off"))
print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="PSG1" & species=="homo sapiens"),
                                genes.to.tree=genes.to.tree,
                                combined.fpkm=combined.fpkm,
                                subtree=c("ENSG00000221826","ENSG00000124469"),
                                trees=trees
                                )+
      geom_tiplab(size=2,color="black",fontface="italic")+xlim(0,0.2)+
      theme(legend.position="none"),
      vp=viewport(layout.pos.col=1,layout.pos.row=1,clip="off"))
print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="CTSL" & species=="homo sapiens"),
                                genes.to.tree=genes.to.tree,
                                combined.fpkm=combined.fpkm,
                                subtree=c("ENSMODG00000002844","ENSMUSG00000055679"),
                                trees=trees)+
      geom_tiplab(size=2,color="black",fontface="italic")+xlim(0,0.8)+
      theme(legend.position="none"),
      vp=viewport(layout.pos.col=2,layout.pos.row=1,clip="off"))
print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="PRSS16" &
                                                    species=="homo sapiens"),
                                genes.to.tree=genes.to.tree,
                                subtree=c(   "ENSG00000112812",
                                          "ENSLAFG00000032211"),
                                combined.fpkm=combined.fpkm,
                                trees=trees)+
      geom_tiplab(size=2,color="black",fontface="italic")+xlim(0,1.2),
      vp=viewport(layout.pos.col=1,layout.pos.row=2,clip="off"))
print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="PAG2" &
                                                    species=="bos taurus"),
                                genes.to.tree=genes.to.tree,
                                combined.fpkm=combined.fpkm,
                                subtree=c("ENSSSCG00000013097",
                                          "ENSBTAG00000046085"),
                                trees=trees) +
      geom_tiplab(size=2,color="black",fontface="italic") +
      xlim(0,0.8)+
      theme(legend.position="none"),
      vp=viewport(layout.pos.col=2,layout.pos.row=2,clip="off"))
popViewport(1)

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \bonobo are shown in Supplemental \cref{tab:all_pan_paniscus_placenta}
with a list of all genes in Supplemental
\cref{tab:all_pan_paniscus_placenta}. Expression in the \bonobo
placenta differs somewhat from that of \human, as evidenced by the
makeup of the top 50 genes (Supplemental \cref{tab:all_pan_paniscus_placenta}). The highest expression
 in \bonobo is %
\Sexpr{gene.expression("pan paniscus","PRG2",complete=TRUE)}, %
%
which is involved in insulin growth factor signaling and angiogenesis
during pregnancy \cite{Weyer.ea2011:Placentalregulationofpeptide}.
\textit{PRG2} is also expressed at relatively high levels (though
comparatively much lower) in the human placenta
(\Sexpr{gene.expression("homo sapiens","PRG2")};
Supplemental \cref{tab:all_homo_sapiens_placenta}). Four \bonobo genes
related to the chorionic somatomammotropins
are also expressed at relatively high levels (\textit{GH1} 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034560",complete=TRUE)}, 
\textit{GH1} 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034212",complete=TRUE)}, 
\textit{GH2}
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000030805",complete=TRUE)}, 
and \textit{GH2}
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034209",complete=TRUE)}, 
%
\cref{tab:all_pan_paniscus_placenta}, \cref{fig:csh1_evolution_tree}).
Placental expression of the genes involved in estrogen and
progesterone production (\textit{CYP19A1}, \textit{HSD3B1}
respectively) show lower expression in \bonobo as compared to \human %
(\Sexpr{gene.expression("pan paniscus","CYP19A1",complete=TRUE)},
\cref{fig:estrogen_evolution_tree}; %
\Sexpr{gene.expression("pan paniscus","HSD3B1",complete=TRUE)},
\cref{fig:progesterone_evolution_tree} and
Supplemental \cref{tab:all_homo_sapiens_placenta}) although the expression of
\textit{HSD3B1} is fairly similar to that of \smonkey %
(\Sexpr{gene.expression("ateles fusciceps","HSD3B1",complete=TRUE)}. %
In addition, \textit{CGA} is expressed in the bonobo placenta %
(\Sexpr{gene.expression("pan paniscus","CGA")}), but at much
lower levels than \human as well (Supplemental \cref{tab:all_homo_sapiens_placenta}).
%
Two Ensembl-annotated beta subunits of the chorionic gonadotropin
genes are expressed in \bonobo, one related to \textit{CGB} and
another related to \textit{CGB7}; both are not highly expressed
(\cref{fig:cga_cgb_evolution_tree}). Among the other top 50 expressed
genes, there is high expression of keratins (e.g., \textit{KRT19},
\textit{KRT18}), heat-shock proteins (e.g., \textit{HSP90B1},
\textit{HSPA1A}), serpins (\textit{SERPINE1}, \textit{SERPINE2},
Supplemental \cref{fig:serpine_evolution_tree}), insulin growth factor
2 (\textit{IGF2}) and one of its binding proteins \textit{IGFBP3}, and
the hormone adrenomedullin (\textit{ADM}).

@@

The \opossum placenta shows expression of genes that either do not
exist in eutherians or are not expressed in the term eutherian placenta
(Supplemental \cref{tab:all_opossum_placenta}). The existence of embryonic forms
of hemoglobin in fetal tissues at term, as evidenced by the expression
of \textit{HBE} and \textit{HBZ} is a feature of marsupials that is
generally due to their very short gestation periods. Embryonic
hemoglobins in marsupials have a lower affinity for oxygen and allow
for a lower rate of oxygen uptake from maternal tissues
\cite{Holland.ea1997:Characterizationofembryonicglobin}. One novel
uncharacterized gene that shows high expression (\textit{LOC10001195}) is not
found in eutherians; Ensembl gene trees reveal that homologs of this
gene exist in other amniotes such as birds and turtles, but are not
present in eutherian mammals. The homolog of this gene in chicken is
found to expressed in reproductive tract
\cite{Froman.ea2006:expressedsequencetaganalysis}, but its function is
unknown. \textit{WFDC2} is a member of the whey acidic protein (WAP)
family and involved in innate immunity. It is expressed in the mammary
gland of the tammar wallaby during pregnancy and early lactation
\cite{Sharp.ea2007:Molecularevolutionofmonotreme}, and present in
relative abundance in \opossum placenta.
